---
title: "Katie - Hobo data"
output: html_notebook
---

Set up **environment** 

```{r, echo=TRUE, results='hide'}
#### Clear  environment 
rm(list = ls())         

#### Required packages
library(tidyverse)        # for data organization and manipulation 
library(lubridate)        # for manipulating time variables
library(purrr) # for joining column onto list of dataframes
library(dplyr)
```

Load the raw data all into one big file but make sure that 

- we only read the important parts (skip = 20, skips the first 20 rows of the raw hobo data)
- save the file names as we need these to link to treatments and sites (.id = "hobo_name" stores the original file names as a new variable)

```{r}
# Get the list of CSV files in the "raw_data" folder
csv_files <- list.files("raw-hobo-data", pattern = "\\.csv$", full.names = TRUE)

# Create an empty list to store the data frames
data_frames_list <- list()

# Loop through each CSV file
for (csv_file in csv_files) {
  # Read the CSV file into a data frame, excluding the first 20 rows
  df_name <- tools::file_path_sans_ext(basename(csv_file))  # Extract the file name without extension
  data_frames_list[[df_name]] <- read.csv(csv_file, skip = 20)
}

# Combine all into one, 
hobo_data <- bind_rows(data_frames_list, .id = "hobo_name")

# bind additonal file with pad number and csv name 

str(hobo_data)


```

Next we need to change the time variable to date format 

```{r}
# Convert the "time" variable to a date-time object
hobo_data$time <- ymd_hms(hobo_data$time)

str(hobo_data)

```


```{r}
# Combine pad ID names with existing dataframe 
#hobo_data <- bind_rows(data_frames_list, .id = "hobo_name")

# bind additional file with pad number and csv name 
Logger_ID_List <- read.csv("Logger_ID_List.csv")
# why does this need ot be .csv

# remove last 2 digits from hobo_names as I made mistake when creating Logger_ID_List.csv 
hobo_data$hobo.names.less2 <- substr(hobo_data$hobo_name, 1, nchar(hobo_data$hobo_name) - 2)


# Perform left join on each dataframe in the list
hobo_data <- left_join(hobo_data, Logger_ID_List, by = "hobo.names.less2")


hobo_data <- hobo_data %>% 
  dplyr::select(c(hobo_name, time, temp, Plate_ID)) %>%
  rename(plate_id = Plate_ID)



#We don't need all of the data, right? Can pick a date range here.

#check 
length(unique(hobo_data$plate_id))




```

```{r}

# Set the sampling date
JUN_1_date <- ymd_hms("2023-06-20 06:00:00") 


# Subset the data using filter
JUN_1 <- hobo_data %>%
  
  #select 2 weeks before sampling date
  filter(time >= JUN_1_date - days(14) & time < JUN_1_date) %>%
  
  # filter for daytime hours 8am - 8pm
  filter(hour(time) >= 6, hour(time) < 20) %>% 
  
  group_by(plate_id) %>%
  summarise(mean_temp = mean(temp), 
            max_temp = max(temp),
            sd_temp = sd(temp)
            
            )

hist(JUN_1$mean_temp, breaks = 20)


# order dataframe
my_order <- c(10,21,7,20,5,17,6,19,2,12,13,11,3,15,8,1,9,23,4,18)

# Reorder the DataFrame
JUN_1_reordered <- JUN_1[match(my_order, JUN_1$plate_id), ]

# Check the reordered DataFrame
print(JUN_1_reordered)

```

```{r}

# Set the sampling date
JUL_1_date <- ymd_hms("2023-07-06 06:00:00") 


# Subset the data using filter
JUL_1 <- hobo_data %>%
  
  #select 2 weeks before sampling date
  filter(time >= JUL_1_date - days(14) & time < JUL_1_date) %>%
  
  # filter for daytime hours 8am - 8pm
  filter(hour(time) >= 6, hour(time) < 20) %>% 
  
  group_by(plate_id) %>%
  summarise(mean_temp = mean(temp), 
            max_temp = max(temp),
            sd_temp = sd(temp)
            
            )

hist(JUL_1$mean_temp, breaks = 20)

JUL_1

JUL_1_reordered <- JUN_1[match(my_order, JUL_1$plate_id), ]

# Check the reordered DataFrame
print(JUL_1_reordered)
```


```{r}

# Set the sampling date
JUL_2_date <- ymd_hms("2023-07-20 06:00:00") 


# Subset the data using filter
JUL_2 <- hobo_data %>%
  
  #select 2 weeks before sampling date
  filter(time >= JUL_2_date - days(14) & time < JUL_2_date) %>%
  
  # filter for daytime hours 8am - 8pm
  filter(hour(time) >= 6, hour(time) < 20) %>% 
  
  group_by(plate_id) %>%
  summarise(mean_temp = mean(temp), 
            max_temp = max(temp),
            sd_temp = sd(temp)
            
            )

hist(JUL_2$mean_temp, breaks = 20)


JUL_2_reordered <- JUL_2[match(my_order, JUL_2$plate_id), ]

# Check the reordered DataFrame
print(JUL_2_reordered)
```

```{r}

# Set the sampling date
AUG_1_date <- ymd_hms("2023-08-01 06:00:00") 


# Subset the data using filter
AUG_1 <- hobo_data %>%
  
  #select 2 weeks before sampling date
  filter(time >= AUG_1_date - days(14) & time < AUG_1_date) %>%
  
  # filter for daytime hours 8am - 8pm
  filter(hour(time) >= 6, hour(time) < 20) %>% 
  
  group_by(plate_id) %>%
  summarise(mean_temp = mean(temp), 
            max_temp = max(temp),
            sd_temp = sd(temp)
            
            )

hist(AUG_1$mean_temp, breaks = 20)


AUG_1_reordered <- AUG_1[match(my_order, AUG_1$plate_id), ]

# Check the reordered DataFrame
print(AUG_1_reordered)
```


```{r}

# Set the sampling date
AUG_2_date <- ymd_hms("2023-08-17 06:00:00") 


# Subset the data using filter
AUG_2 <- hobo_data %>%
  
  #select 2 weeks before sampling date
  filter(time >= AUG_2_date - days(14) & time < AUG_2_date) %>%
  
  # filter for daytime hours 8am - 8pm
  filter(hour(time) >= 6, hour(time) < 20) %>% 
  
  group_by(plate_id) %>%
  summarise(mean_temp = mean(temp), 
            max_temp = max(temp),
            sd_temp = sd(temp)
            
            )

hist(AUG_2$mean_temp, breaks = 20)


AUG_2_reordered <- JUN_1[match(my_order, AUG_2$plate_id), ]

# Check the reordered DataFrame
print(AUG_2_reordered)
```




```{r}

# Set the sampling date
SEP_1_date <- ymd_hms("2023-09-01 06:00:00") 


# Subset the data using filter
SEP_1 <- hobo_data %>%
  
  #select 2 weeks before sampling date
  filter(time >= SEP_1_date - days(14) & time < SEP_1_date) %>%
  
  # filter for daytime hours 8am - 8pm
  filter(hour(time) >= 6, hour(time) < 20) %>% 
  
  group_by(plate_id) %>%
  summarise(mean_temp = mean(temp), 
            max_temp = max(temp),
            sd_temp = sd(temp)
            
            )

hist(SEP_1$mean_temp, breaks = 20)


SEP_1_reordered <- SEP_1[match(my_order, SEP_1$plate_id), ]

# Check the reordered DataFrame
print(SEP_1_reordered)
```


```{r}

# Set the sampling date
SEP_2_date <- ymd_hms("2023-09-15 06:00:00") 


# Subset the data using filter
SEP_2 <- hobo_data %>%
  
  #select 2 weeks before sampling date
  filter(time >= SEP_1_date - days(14) & time < SEP_1_date) %>%
  
  # filter for daytime hours 8am - 8pm
  filter(hour(time) >= 6, hour(time) < 20) %>% 
  
  group_by(plate_id) %>%
  summarise(mean_temp = mean(temp), 
            max_temp = max(temp),
            sd_temp = sd(temp)
            
            )

hist(SEP_2$mean_temp, breaks = 20)

SEP_2_reordered <- SEP_2[match(my_order, SEP_2$plate_id), ]

# Check the reordered DataFrame
print(SEP_2_reordered)
```
```{r}
# Assuming hobo_data is your dataset and you've already done the previous steps

# Define the pad ID numbers for each group
control_logs <- c('10','7','5','6','2','12','11','3','8','1','9','4') 
warm_logs <- c('21','20', '17','19','13','15','23','18') # Replace with actual plate_id values for Group 2

# Categorize each row based on its plate_id

hobo_data <- hobo_data %>%
  mutate(group = case_when(
    plate_id %in% warm_logs ~ 'Warm Pads',
    plate_id %in% control_logs ~ 'Control Pads'
  ))

# hobo_data will have a new column 'group' categorizing each row



```
### this chunk needs help
```{r}


hobo_data$date <- as.Date(hobo_data$time)

hobo_data$Time <- round_date(hobo_data$time, "hour")


# Calculating the mean for each group at each date
average_hobo_data <- hobo_data %>%
  group_by(Time, group) %>%
  summarize(mean_temp = mean(temp, na.rm = TRUE), .groups = 'drop')

# Calculating the max for each group at each date
max_hobo_data <- hobo_data %>%
  group_by(Time, group) %>%
  summarize(max_temp = max(temp, na.rm = TRUE), .groups = 'drop')


average_hobo_data$Time <- as.POSIXct(average_hobo_data$Time, format="%Y-%m-%d %H:%M:%S")

max_hobo_data$Time  <- as.POSIXct(max_hobo_data$Time, format="%Y-%m-%d %H:%M:%S")

#sort out averages

#tide$date <- as.Date(tide$date)
```

```{r}
# Load the data
tide <- read.table("Brighton_tide_heights.txt", header = TRUE, sep = "", skip = 11)


names(tide) <- c('Row', 'Date', 'Time', 'Height', 'Residual')

tide <- tide %>%
  mutate(DateTime = as.POSIXct(paste(Date, Time), format = "%Y/%m/%d %H:%M:%S"))

# Now filter for a specific date range, for example, March 1, 2023, to March 31, 2023
filtered_tide <- tide %>%
  filter(DateTime >= as.POSIXct("2023-06-20") & DateTime <= as.POSIXct("2023-09-15"))


# Filter out rows where Height is not finite
filtered_tide_clean <- filtered_tide %>% 
  filter(is.finite(Height))


# Convert Time to POSIXct in average_hobo_data
hobo_data$Time <- as.POSIXct(hobo_data$Time, format = "%Y-%m-%d %H:%M:%S")
# Convert Time to POSIXct in filtered_tide
filtered_tide$Time <- as.POSIXct(filtered_tide$DateTime, format = "%Y-%m-%d %H:%M:%S")



filtered_tide$Height <- gsub("M", "", filtered_tide$Height)  # Remove 'M' from the Height values

# Convert Height to numeric
filtered_tide$Height <- as.numeric(filtered_tide$Height)

# Check for any Inf, -Inf, or NaN values in the Height column
sum(!is.finite(filtered_tide$Height))

# Check for any conversion warnings or errors
ggplot() +
  geom_line(data = filtered_tide, aes(x = Time, y = Height), colour = "lightblue")



```



```{r}
# Assuming max_hobo_data exists and is correct
ggplot() +
  geom_point(data = max_hobo_data %>% filter(group == "Warm Pads"), aes(x = Time, y = max_temp, color = "Warm Pads")) +
  geom_point(data = max_hobo_data %>% filter(group == "Control Pads"), aes(x = Time, y = max_temp, color = "Control Pads")) +
  scale_color_manual(values = c("Warm Pads" = "red", "Control Pads" = "blue")) +
  geom_line(data = max_hobo_data, aes(x = Time, y = max_temp, group = group, color = group)) +
  labs(title = "Temperature Trends",
       x = "Date",
       y = "Max Temperature",
       color = "Group") +
  theme_minimal() +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%b %d", limits = as.POSIXct(c("2023-08-19", "2023-08-23")))

# Note: Corrected the date in the upper limit from "2023-08-123" to "2023-08-23"


```



```{r}
ggplot() +
  geom_point(data = average_hobo_data %>% filter(group == "Warm Pads"), aes(x = Time, y = mean_temp, color = group)) +
  geom_line(data = average_hobo_data %>% filter(group == "Warm Pads"), aes(x = Time, y = mean_temp, group = group), color = "red") +
  geom_point(data = average_hobo_data %>% filter(group == "Control Pads"), aes(x = Time, y = mean_temp, color = group)) +
  geom_line(data = average_hobo_data %>% filter(group == "Control Pads"), aes(x = Time, y = mean_temp, group = group), color = "blue") +
  scale_color_manual(values = c("Warm Pads" = "red", "Control Pads" = "blue")) +
  labs(title = "Temperature Trends", x = "Date", y = "Temperature", color = "Group") +
  theme_minimal() +
  scale_x_datetime(limits = as.POSIXct(c("2023-08-20 00:00:00", "2023-08-25 00:00:00")))+
  ylim(10,35)

#what does this waerning mean!?
#make sure low tide isn't at night!
#can chekc this against tidal data

```
```{r}
#temperature doesn't match up with average_hobo_data



tide_plot <- ggplot() +
  geom_point(data = average_hobo_data %>% filter(group == "Warm Pads"), aes(x = Time, y = mean_temp, color = group)) +
  geom_line(data = average_hobo_data %>% filter(group == "Warm Pads"), aes(x = Time, y = mean_temp), color = "red") +
  geom_point(data = average_hobo_data %>% filter(group == "Control Pads"), aes(x = Time, y = mean_temp, color = group)) +
  geom_line(data = average_hobo_data %>% filter(group == "Control Pads"), aes(x = Time, y = mean_temp), color = "blue") +
  geom_area(data = filtered_tide, aes(x = Time, y = Height, fill = "Tide"), colour = "lightblue", alpha = 0.5) + # Fill under the tide line
 # geom_ribbon(data = hobo_se, aes(x = Time, ymin = Lower, ymax = Upper, fill = group), alpha = 0.2) +
  scale_fill_manual(values = c("Warm Pads" = "red", "Control Pads" = "blue", "Tide" = "lightblue")) +
  scale_color_manual(values = c("Warm Pads" = "red", "Control Pads" = "blue")) +
  scale_y_continuous("Temperature Scale",
                     sec.axis = sec_axis(~ . * max(filtered_tide$Height) / max(hobo_data$temp), name = "Tide Height Scale")) +
  labs(x = "Date") +
   scale_x_datetime(date_labels = "%Y-%m-%d %H:%M", date_breaks = "1 day", limits = as.POSIXct(c("2023-08-07 10:00:00", "2023-08-09 10:00:00")))+
  theme_minimal()

tide_plot

```


```{r}

hobo_se <- hobo_data %>%
  group_by(Time, group) %>%
  summarise(
    Avg = mean(temp),
    SE = sd(temp) / sqrt(n()),
    Lower = Avg - SE,  # Assuming 1 SE for the demonstration
    Upper = Avg + SE
  ) %>%
  ungroup()


#try with straigth hobot data

tide_plot_2 <- ggplot() +
  geom_point(data = hobo_data %>% filter(group == "Warm Pads"), aes(x = Time, y = temp, color = group), size = 1, alpha = 0.2) +
  geom_line(data = hobo_data %>% filter(group == "Warm Pads"), aes(x = Time, y = temp, group = group),  color = "red", alpha = 0.8) +
  geom_point(data = hobo_data %>% filter(group == "Control Pads"), aes(x = Time, y = temp, color = group), size = 1, alpha = 0.1) +
  geom_line(data = hobo_data %>% filter(group == "Control Pads"), aes(x = Time, y = temp, group = group), color = "blue", alpha = 0.8) +
  scale_color_manual(values = c("Warm Pads" = "red", "Control Pads" = "blue")) +
geom_area(data = filtered_tide, aes(x = Time, y = Height, fill = "Tide"), colour = "lightblue", alpha = 0.5) + # Fill under the tide line
  scale_fill_manual(values = c("Tide" = "lightblue"))+
  scale_y_continuous("Temperature Scale",
                     sec.axis = sec_axis(~ . * max(filtered_tide$Height) / max(hobo_data$temp), name = "Tide Height Scale")) +
  labs(x = "Date") +
   scale_x_datetime(date_labels = "%Y-%m-%d %H:%M", date_breaks = "1 day", limits = as.POSIXct(c("2023-09-07 00:00:00", "2023-09-09 00:00:00")))+
  theme_minimal()+
   geom_ribbon(data = hobo_se, aes(x = Time, ymin = Lower, ymax = Upper, group = group, fill = group), alpha = 0.2)+
  scale_color_manual(values = c("Warm Pads" = "red", "Control Pads" = "blue")) # Add this line for shaded areas

tide_plot_2

```


```{r}

tide_plot_3 <- ggplot() +
  geom_point(data = hobo_data %>% filter(group == "Warm Pads"), aes(x = Time, y = temp, color = group), size = 1, alpha = 0.4) +
  #geom_line(data = hobo_data %>% filter(group == "Warm Pads"), aes(x = Time, y = temp, group = group), color = "red", alpha = 0.8) +
  geom_point(data = hobo_data %>% filter(group == "Control Pads"), aes(x = Time, y = temp, color = group), size = 1, alpha = 0.4) +
  #geom_line(data = hobo_data %>% filter(group == "Control Pads"), aes(x = Time, y = temp, group = group), color = "blue", alpha = 0.8) +
  geom_area(data = filtered_tide, aes(x = Time, y = Height, fill = "Tide"), color = "lightblue", alpha = 0.5) +
  geom_ribbon(data = hobo_se, aes(x = Time, ymin = Lower, ymax = Upper, fill = group), alpha = 0.4) +
  scale_fill_manual(values = c("Warm Pads" = "red", "Control Pads" = "blue", "Tide" = "lightblue")) +
  scale_color_manual(values = c("Warm Pads" = "red", "Control Pads" = "blue")) +
  labs(x = "Date", y = "Temperature Scale", title = "Temperature and Tide Trends") +
  scale_x_datetime(date_labels = "%Y-%m-%d %H:%M", date_breaks = "1 day", limits = as.POSIXct(c("2023-09-07 00:00:00", "2023-09-09 00:00:00"))) +
  theme_minimal()




tide_plot_3


#ggsave("tide_plot.pdf", tide_plot, width = 11, height = 8)

#why is there no difference between the temp treatments?

```


```{r}
#add air
 # geom_line(data = brighton_temp, aes(x = date, y = temp), color = "gray", linetype = "dashed") +
```


Important next step is to match up the hobo file names with the site information 


Then we can get summary data (like average temperature over a 2 week period?)


Then we can merge the summary data with the main dataframe 
